---
title: "Class 12: Ensembl"
author: "Thomas Bailey"
format: pdf
---

## Section 1. Propotion og G/G in a population 

Downloaded a CSV file from Ensemble < link here 

Here we read this CSV file 


```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) *100
```

Now let's look at a different population. I pickd the GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(gbr)
```

Find the proportion of G|G

```{r}
table(gbr$Genotype..forward.strand.)/nrow(gbr) * 100 
```

This varient that is associated with childhood asthma is more frequent in the GBR population than the MXL population. 

Lets now dig into this further. 

## Section 4: Population Scale Analysis
One smaple is obviously not enough to know what is happening in a population you are interested in assessing genetic differences on a population scale. 

How many samples do we have? 
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```


```{r}
nrow(expr)
```
> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
table(expr$geno)
```
```{r}
library(ggplot2) 

```


Lets make a boxplot

```{r}
ggplot(expr) + aes(x=geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```


```{r}
summary(expr)
```
> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

based on the boxplot it seems as though the SNP does effect the expression of ORMDL3. 




