---
title: "R Functions"
author: "Thomas Bailey"
format: pdf
---

today we will get more exposure to functions in R. We call functions to do all our work in R.

##A first silly function

Note that arguments 2 and 3 have default values (because we set y=0 and z=0) so we don't have to supply them when we call our functions.


```{r}
add <- function(x,y=0,z=0) {
  x + y + z
}
```

```{r}
add(1,1)
add(1,c(10,100))
```

Can I just use this 

```{r}
add(100)
```

```{r}
add(100,10,1)
```
## A second more fun function

let's write a function that generates random nucleotide sequences.

We can make use of the in-built `sample()` function in R to help us here. 


```{r}
sample(x=1:10, size=9)
```

```{r}
sample(x=1:10, size=11, replace = TRUE)
```

>Q. Can you use `sample()` to generate a random nucleotide sequence of length 5. 
##yes you can set x = to the bases the set size to 5 and put true on replace

```{r}
sample(x=c("A", "C", "G", "T"), size= 5, replace= T)
```

>Q. Write a function `generate_dna()` that makes a nucleotide sequence of a user specified length 

every function in R has at least 3 things: 

-a **name** (in our case "generate_dna")
-one ore more **input arguments** (the "length" of sequence we want)
-a **body** (that does the work)

```{r}
generate_dna <- function(length=5) {
  bases <- c("A", "C", "G", "T")
  sample(bases, size=length, replace=T)
  
}
  
```
    
    
```{r}
    generate_dna(17)
```
    
    
    
>Can you write a `generate_protein()` function that returns amino acid sequence of a user requested length?. 
  
I want my outpost of this function not to b a vector with one amino acid per element but rather a one element single string.

```{r}

bases<- c("A", "G", "C", "T")

paste(bases,collapse=" ")
```


```{r}
generate_protein <- function(length=5){
  
aa <- bio3d::aa.table$aa1[1:20]
s <- sample(aa, size=length, replace=T)
paste(s, collapse=" ")
}
```


We can use the 

```{r}
ans <- sapply(6:12,generate_protein)
```

```{r}
cat(paste(">ID", 6:12, sep="", "\n", ans, "\n"))
```

>Q. Are any of these sequences unique in nature - i.e. never found in nature. We can search "refseq-protein"  and look for 100% Ide and 100% Coverage.

In looking up in blastp we found many similar sequences but with high E-Values suggesting that even though they might say they have 100% Ide and 100% Coverage that they are still not naturally found out in nature so they are unique sequences. 



