---
title: "Comparative structure analysis"
author: "Thomas Bailey"
format: pdf
---

Here we run through a complete "pipeline" of structure analysis that begins with a singe sequence  identifier and ends with a PCA analysis. 

```{r}
library(bio3d)
```

## Step 1. retrive a sequence for the protein we are interested in. We will take ADK "1ake_A"

```{r}
id <- "1ake_A" 
aa <- get.seq(id)
aa
```
## Step 2. Blast search 

Run a BLAST search of the PDB for all related sequences to our input `aa`

```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```
## Step 3. Download all structures 
These are our "top hits" i.e. all the structures in the PDB  database related to our input sequence

```{r}
hits$pdb.id
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

## step 4. Align and superpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```


## Step 5. PCA

Let's use our old friend PCA to make sense of these confussing, complicated structure relationships


```{r}
pc <- pca(pdbs)
plot(pc)
```

Let's make a trajectory (or movie) of the main conformational changes capptured by PC1. We will use the `mktrj()` function for this... 

```{r}
mktrj(pc, file="pca_result.pdb")
```


Back of the envelope comparison of the PDB size to UniProt 

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot * 100
```




